Pancreatic endocrinogenesis

### 2024-08-07 nbs/tutorials/pancreas/results/pancreas.qmd
docs(tutorials): add more detailed gene selection data
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

execution_id = "pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a"

@@

(
    model2_postprocessing_inputs,
    model2_postprocessing_outputs,
) = get_remote_task_results(
    execution_id=execution_id,
    task_id="f26c1pjy-0-dn1-0-dn6",
)
```

@@

```{python}
# | label: model1-extract-gene-selection
from pyrovelocity.analysis.analyze import pareto_frontier_genes

volcano_data = posterior_samples["gene_ranking"]
number_of_marker_genes = min(
    max(int(len(volcano_data) * 0.1), 4), 6, len(volcano_data)
)
putative_marker_genes = pareto_frontier_genes(
    volcano_data, number_of_marker_genes
)
```

@@

```{python}
# | label: model1-generate-gene-selection-summary-plot
# | output: false
from pyrovelocity.plots import plot_gene_selection_summary

vector_field_basis = model1_postprocessing_inputs.preprocess_data_args.vector_field_basis
cell_state = model1_postprocessing_inputs.preprocess_data_args.cell_state

plot_gene_selection_summary(
    adata=adata,
    posterior_samples=posterior_samples,
    basis=vector_field_basis,
    cell_state=cell_state,
    plot_name="gene_selection_summary_plot.pdf",
    selected_genes=putative_marker_genes,
    show_marginal_histograms=False,
)
```

```{python}
# | label: model1-show-gene-selection-summary-plot
# | code-fold: true
# | output: true
display(Image(filename=f"gene_selection_summary_plot.pdf.png"))
```

@@

@@

```{python}
# | label: model2-extract-gene-selection
from pyrovelocity.analysis.analyze import pareto_frontier_genes

volcano_data = posterior_samples["gene_ranking"]
number_of_marker_genes = min(
    max(int(len(volcano_data) * 0.1), 4), 6, len(volcano_data)
)
putative_marker_genes = pareto_frontier_genes(
    volcano_data, number_of_marker_genes
)
```

@@

```{python}
# | label: model2-generate-gene-selection-summary-plot
# | output: false
from pyrovelocity.plots import plot_gene_selection_summary

vector_field_basis = model2_postprocessing_inputs.preprocess_data_args.vector_field_basis
cell_state = model2_postprocessing_inputs.preprocess_data_args.cell_state

plot_gene_selection_summary(
    adata=adata,
    posterior_samples=posterior_samples,
    basis=vector_field_basis,
    cell_state=cell_state,
    plot_name="gene_selection_summary_plot.pdf",
    selected_genes=putative_marker_genes,
    show_marginal_histograms=False,
)
```

```{python}
# | label: model2-show-gene-selection-summary-plot
# | code-fold: true
# | output: true
display(Image(filename=f"gene_selection_summary_plot.pdf.png"))
```

>>>>

# execution_id = "pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a"  # downsampled
execution_id = (
    "pyrovelocity-595-gsmc-feaa936-dev-vf0-264734a284ac493fa39"  # complete
)

@@

(
    model2_postprocessing_inputs,
    model2_postprocessing_outputs,
) = get_remote_task_results(
    execution_id=execution_id,
    task_id="f26c1pjy-0-dn1-0-dn7",
)
```

@@

We decompose the correlation of each gene's expression level with the
cell time estimates into eight equally-spaced bins and select the top 3% of genes with respect to the mean absolute error between the predictive sample mean and
the observed expression level.

```{python}
# | label: model1-extract-gene-selection
from pyrovelocity.analysis.analyze import top_mae_genes

volcano_data = posterior_samples["gene_ranking"]
number_of_marker_genes = min(
    max(int(len(volcano_data) * 0.1), 4), 6, len(volcano_data)
)
putative_marker_genes = top_mae_genes(
    volcano_data,
    mae_top_percentile=3,
    min_genes_per_bin=3,
)
```

@@

We generate several summary plots below for two distinct model classes 
in order to support comparative model evaluation, which is among few
options for selecting effective models that are not designed to be
microscopically valid.

Here we plot the gene selection statistics highlighting genes of interest
based upon them.

```{python}
# | label: model1-generate-gene-selection-plot
# | output: false
from pyrovelocity.plots import plot_gene_ranking

vector_field_basis = (
    model1_postprocessing_inputs.preprocess_data_args.vector_field_basis
)
cell_state = model1_postprocessing_inputs.preprocess_data_args.cell_state

plot_gene_ranking(
    posterior_samples=[posterior_samples],
    adata=[adata],
    selected_genes=putative_marker_genes,
    time_correlation_with="st",
    show_marginal_histograms=True,
    save_volcano_plot=True,
    volcano_plot_path="model1_volcano_plot.pdf",
)
```
```{python}
# | label: model1-show-gene-selection-plot
# | code-fold: true
# | output: true
display(Image(filename=f"model1_volcano_plot.pdf.png"))
```

We can attempt to attribute the statistics to underlying features plotting 
phase portraits, predictive expression, and log-transformed expression of
the proposed marker genes

```{python}
# | label: model1-generate-rainbow-plot
# | output: false
from pyrovelocity.plots import rainbowplot

rainbowplot(
    volcano_data=volcano_data,
    adata=adata,
    posterior_samples=posterior_samples,
    genes=putative_marker_genes,
    data=["st", "ut"],
    basis=vector_field_basis,
    cell_state=cell_state,
    save_plot=True,
    rainbow_plot_path="model1_rainbow_plot.pdf",
)
```

```{python}
# | label: model1-show-rainbow-plot
# | code-fold: true
# | output: true
display(Image(filename=f"model1_rainbow_plot.pdf.png"))
```

@@

We provide a comparable analysis for the second model below.

@@

```{python}
# | label: model2-extract-gene-selection
from pyrovelocity.analysis.analyze import top_mae_genes

volcano_data = posterior_samples["gene_ranking"]
number_of_marker_genes = min(
    max(int(len(volcano_data) * 0.1), 4), 6, len(volcano_data)
)
putative_marker_genes = top_mae_genes(
    volcano_data,
    mae_top_percentile=3,
    min_genes_per_bin=3,
)
```

@@

```{python}
# | label: model2-generate-gene-selection-plot
# | output: false
from pyrovelocity.plots import plot_gene_ranking

vector_field_basis = (
    model2_postprocessing_inputs.preprocess_data_args.vector_field_basis
)
cell_state = model2_postprocessing_inputs.preprocess_data_args.cell_state

plot_gene_ranking(
    posterior_samples=[posterior_samples],
    adata=[adata],
    selected_genes=putative_marker_genes,
    time_correlation_with="st",
    show_marginal_histograms=True,
    save_volcano_plot=True,
    volcano_plot_path="model2_volcano_plot.pdf",
)
```
```{python}
# | label: model2-show-gene-selection-plot
# | code-fold: true
# | output: true
display(Image(filename=f"model2_volcano_plot.pdf.png"))
```

Plot phase portraits, predictive expression, and log-transformed expression of
all proposed marker genes

```{python}
# | label: model2-generate-rainbow-plot
# | output: false
from pyrovelocity.plots import rainbowplot

rainbowplot(
    volcano_data=volcano_data,
    adata=adata,
    posterior_samples=posterior_samples,
    genes=putative_marker_genes,
    data=["st", "ut"],
    basis=vector_field_basis,
    cell_state=cell_state,
    save_plot=True,
    rainbow_plot_path="model2_rainbow_plot.pdf",
)
```

```{python}
# | label: model2-show-rainbow-plot
# | code-fold: true
# | output: true
display(Image(filename=f"model2_rainbow_plot.pdf.png"))
```
